Linear analysis of the vertical shear instability in astrophysical disks

### 2015-02-10 paper.tex
writing introduction

However, MHD turbulence may not be applicable for all 
accretion disks. For example, there may be regions in protoplanetary
disks --- called dead zones ---  which do not support the MRI 
\citep{gammie96,landry13}. In fact, recent state-of-the-art

%in this respect purely hydrodynamical turbulence should be

%stuff about vsi

%recent calculations (nelson, mcnally)

>>>>

However, MHD turbulence may not be applicable for all  
accretion disks. For example, there may be regions in protoplanetary  
disks (PPDs) which do not support the MRI 
\citep{gammie96,landry13}. In fact, recent  numerical simulations
suggest portions of PPDs remain laminar \citep[e.g.][and references
therein]{lesur14,bai15,gressel15}. In this respect, other mechanisms 
to angular momentum transport in PPDs remain of interest
\citep{armitage10}.

%recent work on vsi 
One possibiliy is purely hydrodynamic turbulence through the vertical
shear instability (VSI). The VSI is an axisymmetric linear instability
in three-dimensional (3D) disks in which the rotation frequency
$\Omega$ depends on the height from the disk midplane. This
height-dependence, or vertical shear, enables perturbations with
large vertical lengthscales (relative to radial lengthscales) to
become unstable. The VSI is the disk analog of the
Goldreich-Schubert-Fricke instability
\citep[GSF,]{goldreich67,fricke68} in differentially rotating stars.   
\cite{umurhan13} gives a simple explanation for this 
instability: an exchange of fluid parcels while conserving their
angular momenta individually, may result in a new configuration of
lower energy, if the exchange occurs over a sufficiently large
vertical displacement.

The VSI has

%history

recently been demonstrated through 
hydrodynamical simulations \citep{nelson13}

%and kley simulations

%vsi background

%questions for this work

### 2015-02-10 summary.tex
writing introduction

We briefly studied vertically non-isothermal disks using our linear 
code. We confirm the finding by \cite{nelson13} that in the absence
of vertical entropy gradients, the VSI operates and is insensitive to
thermal relaxation. When a stable stratification is introduced, the
rapid reduction in VSI growth rates by a finite thermal relaxation
timescale is qualitatively similar to that observed for vertically
isothermal disks. Based on numerical results, we conjecture that the
critical thermal relaxation timescale is given by
$\beta_\mathrm{crit}=\epsilon|q|/(\gamma-\Gamma)$ in verticlaly
non-isothermal disks.

@@

\subsection{Caveats and outlooks} 
We highlight below some improvements to the present  
linear models that should be pursued:

@@

\emph{Additional physics.} Depending on where the VSI operates in a
disk, additional effects may become important. For example, in the
outer parts of a protoplanetary disk, where we estimate the VSI may
operate, disk self-gravity (SG) may play a role. Specifically, the vertical
shear profile depends on the vertical density stratification
(Eq. \ref{vertical_shear}), which wil be enhanced by vertical SG,
which will enhance the VSI. Inclusion of the Poisson equation should
therefore be considered.

@@

\emph{Geometry.} An important caveat

>>>>

We briefly studied vertically non-isothermal disks using our linear 
code. We confirm the finding by \cite{nelson13} that in the absence
of vertical entropy gradients, the VSI operates and is insensitive to
thermal relaxation. When a stable stratification is introduced, the
rapid reduction in VSI growth rates by a finite thermal relaxation
timescale is qualitatively similar to that observed for vertically
isothermal disks. Based on numerical results, we conjecture that the
critical thermal relaxation timescale is given by
$\beta_\mathrm{crit}=\epsilon|q|/(\gamma-\Gamma)$ in vertically 
non-isothermal disks.

@@

\subsection{Caveats and outlooks} 
We highlight below some improvements to the present  
linear models in future works:

@@

\emph{Additional physics.} Depending on where the VSI operates in a
disk, additional effects may become important. For example, in the
outer parts of a protoplanetary disk, where we estimate the VSI 
operates, disk self-gravity (SG) may play a role. Specifically, the vertical
shear profile depends on the vertical density stratification
(Eq. \ref{vertical_shear}) which can be enhanced by vertical SG. This
may promote the VSI. Inclusion of the Poisson equation should
therefore be considered.

@@

\emph{Radial dependence.} We have adopted the radially-local approximation 
throughout this paper. This is based on previous works, including
nonlinear simulations, which show that the VSI is relevant for
perturbations with small radial lengthscales. Although this simplifies
the linear problem to ordinary differential equations, this approach
introduces the wavenumber $k_x$ which is a free parameter. The effect
of radial boundary conditions cannot be explored. Future
generalizations of the linear problem should retain the radial
dependence properly and solve the two-dimensional $(r,z)$ partial 
differential equation.
